Extended Data

### 2022-06-04 doc/manuscript/Dow_manuscript_Nature.Rmd
update code version numbers and URLs -- still need ARSTAN v number
#148

<!--Hypothesis testing -->
Correlation between the dendrometer band-derived growth parameters (*DOY~25~*, *DOY~50~*, *DOY~75~*, *L~pgs~*, *g~max~*, and *∆DBH*, Fig. `r schematic_summary`) and _CTW_ *T~max~* (at weekly or monthly resolution, as described above) were assessed using a linear mixed model in a hierarchical Bayesian framework.
Analyses were run for both *T~max~* and *T~min~*, with qualitatively similar results, but we present only results for *T~max~*, which had overall stronger correlation with growth parameters. 
Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean *T~max~* (or *T~min~*) during the _CTW_, along with random effects of species and of individual tree. 
We ran separate models for each site, and for the response of all growth parameters to *T~max~* (or *T~min~*). 
This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020;@standevelopmentteam_stan_2019]. 
In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
<!--
More information on rstanarm's default weakly informative priors:
https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
-->

@@

For each species-site combination, we converted tree-ring records into the dimensionless _RWI_ to emphasize interannual variability associated with climate[@cook_methods_1990].
A 2/3rds _n_ spline was applied to each core using ARSTAN to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985;@cook_methods_1990].
An adaptive power transformation, a process that also stabilizes the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 
Low series replication, often in the earliest portions of a chronology collection, can also inflate the variance of tree-ring records[@jones_estimating_1997].
The 1/3rds spline method was chosen when replication in the inner portion of each chronology (the earliest ca. 30–50 yr of each record depending on full chronology length) was less than three trees. 
When replication was greater than n = 3 trees, we used the average correlation between raw ring-width series (rbar) method.
The robust biweight mean chronology (_RWI_) for each species-site combination was calculated from the ring-width indices following variance stabilization[@cook_time_1985].
We defined chronology start year (`r EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where ≥80% of the population signal was captured in the chronology.

@@

# Code Availability Statement {-}
Data were analysed in the open source statistical software R (version 4.0).
We used packages *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html),  *dplR* v.1.0.2, and *bootRes* v1.2.4, *rstanarm* v.2.21.1, and functions from *Rdendrom* v.0.1.0 (https://github.com/seanmcm/RDendrom/).
We used *climpact* software v.1.2.8 (see www.climpact-sci.org).
All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: *[TBD]*).

>>>>

<!--Hypothesis testing -->
Correlation between the dendrometer band-derived growth parameters (*DOY~25~*, *DOY~50~*, *DOY~75~*, *L~pgs~*, *g~max~*, and *∆DBH*, Fig. `r schematic_summary`) and _CTW_ *T~max~* (at weekly or monthly resolution, as described above) were assessed using a linear mixed model in a hierarchical Bayesian framework.
Analyses were run for both *T~max~* and *T~min~*, with qualitatively similar results, but we present only results for *T~max~*, which had overall stronger correlation with growth parameters. 
Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean *T~max~* (or *T~min~*) during the _CTW_, along with random effects of species and of individual tree. 
We ran separate models for each site, and for the response of all growth parameters to *T~max~* (or *T~min~*). 
This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` version 2.21.3 R interface to the Stan programming language [@gabry_rstanarm_2020;@standevelopmentteam_stan_2019]. 
In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
<!--
More information on rstanarm's default weakly informative priors:
https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
-->

@@

For each species-site combination, we converted tree-ring records into the dimensionless _RWI_ to emphasize interannual variability associated with climate[@cook_methods_1990].
A 2/3rds _n_ spline was applied to each core using ARSTAN **v.###** to produce standardized ring-width series; _n_ is the number of years in each series[@cook_time_1985;@cook_methods_1990].
An adaptive power transformation, a process that also stabilizes the variance over time[@cook_calculating_1997], was used to minimize the influence of outliers in all series. 
Low series replication, often in the earliest portions of a chronology collection, can also inflate the variance of tree-ring records[@jones_estimating_1997].
The 1/3rds spline method was chosen when replication in the inner portion of each chronology (the earliest ca. 30–50 yr of each record depending on full chronology length) was less than three trees. 
When replication was greater than n = 3 trees, we used the average correlation between raw ring-width series (rbar) method.
The robust biweight mean chronology (_RWI_) for each species-site combination was calculated from the ring-width indices following variance stabilization[@cook_time_1985].
We defined chronology start year (`r EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where ≥80% of the population signal was captured in the chronology.

@@

# Code Availability Statement {-}
<!--this statement to be replicated in "nr-reporting-summary.pdf"-->
Data were analysed in the open source statistical software R (version 4.0).
We used packages *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html),  *dplR* v.1.0.2, and *bootRes* v1.2.4, *rstanarm* v.2.21.1, and functions from *Rdendrom* v.0.1.0 (https://github.com/seanmcm/RDendrom/).
We used *climpact* software v.1.2.8 (see www.climpact-sci.org).
Mixed effects models were run with mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` version 2.21.3 R interface to the Stan programming language (source code available at https://github.com/EcoClimLab/growth_phenology#steps-to-replicate-the-analysis). 
Tree-ring chronologies were developed using the program ARSTAN **v.###** (https://www.geog.cam.ac.uk/research/projects/dendrosoftware/).
All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: *[TBD]*).
